---
title: "R4ds ch. 3 Data visualization"
output: html_notebook
---

This script follows the book R4ds by Hadley wickham. Source: https://r4ds.had.co.nz/

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

3.2 First steps
3.2.1 The mpg data frame

plot mpg dataframe fuel efficiency on highway vs car engine size 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

create ggplot2 template

ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

3.2.4 Exercises

Q. Run ggplot(data = mpg). What do you see?
A. an empty plot  

Q. How many rows are in mpg? How many columns?
A. 234 rows and 11 columns  

Q. What does the drv variable describe? Read the help for ?mpg to find out.
A. drv stands for type of drive: f = front wheel, 3 = rear wheel drive, 4 = 4wd


Make a scatterplot of hwy vs cyl.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

Q. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))
```




3.3 aesthetic mapping

assign a unique color to each car class
```{r}
ggplot( data = mpg) + 
  geom_point(mapping = aes( x = displ, y = hwy, color = class ))
```

assign different point size for each car class
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

same plot with transparent point based on class
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes( x = displ, y = hwy, alpha = class))
```

same plot with different shape for each car class

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

scatterplot between fuel efficiency and highway mileages 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes( x = displ, y = hwy), color = 'blue')
```

3.3.1 exercises

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y =  hwy, color = displ < 5, shape = drv))
```

split points by fuel efficiency treshold 

```{r}
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y =  hwy, color = displ < 5, shape = drv))
```



3.5 Facets

create multiple plots for different classes of a single variable with facet  

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) # facet wrap needs a discrete variable
```

facet a plot on combination of 2 variables. 

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl) # facet grid should contain two variables
```

facet grid: scatter plot between engine displacement (displ), miles per gallon (hwy), sorted by number of cylinders 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid( . ~ cyl) # use . if you don't want to facet rows or columns
```

3.5.1 Exercises

1) facet wrap using a continious variable splits plots by value
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ cty)
```

2)

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl))
```

3) . does 

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
```


4) 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(. ~ class, nrow = 2)
```

3.6 Geometric objects

compare geoms. First plot a scatterplot
```{r}
ggplot( data = mpg) + 
  geom_point(aes(x = displ, y = hwy))
```

then a smoothed geom scatterplot
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x  =  displ, y = hwy))
```

separate cars into three lines based on drv value. 
```{r}
ggplot( data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv  )) # set linetype for different wheel drive (f = front wheel, 4 = four wheel, r = rear wheel)
```


now overlay smoothed lines and points and separate by drive class (4wheel,rear,front)
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes( x = displ, y = hwy, color = drv ,linetype = drv)) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_smooth(mapping = aes( x = displ, y = hwy, group = drv))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, color = drv),
                            show.legend = FALSE)
```


now display multiple geoms in the same plots
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes( x = displ, y = hwy)) + # add scatter plot
  geom_smooth(mapping = aes(x = displ, y = hwy)) # add smooth line 
```


now recreate the same plot without code duplication

```{r}
ggplot(data = mpg, mapping = aes( x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```

mappings for a geom will change the layout only for that layout. Use global mappings for all geoms
```{r}
ggplot(data = mpg, mapping = aes( x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
               geom_smooth()
```

       
now specify different data for each geom layer. Smoothed geom shows a subset of auto class subcompact

```{r}
ggplot(data = mpg, mapping = aes( x = displ, y = hwy)) +
  geom_point(mapping = aes( color = class)) +
  geom_smooth(data = filter(mpg , class == "subcompact"),se = F) # se = confidence interval
```

3.6.1 exercises 
line = geom_line
histogram = geom_hist
boxplot = geom_boxplot()
area chart = geom_area()

play with the show.legend option. This will add the legend for the specified geom.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point(show.legend = T) + 
  geom_smooth(se = T, show.legend = F)
```


3.6.6
```{r}
ggplot( data = mpg, mapping = aes( x = displ, y = hwy)) + # plot scatterplot with smoothed line  
  geom_point()+ 
  geom_smooth(se = F) # remove confidence interval
```

```{r}
ggplot( data = mpg,aes( x = displ, y = hwy) ) + # plot scatterplot and smooth lines 
  geom_point() +
  geom_smooth(mapping = aes(group = drv)  , se = F) # add smooth line for each drv group

ggplot(data = mpg, mapping = aes( x = displ, y = hwy, color = drv)) + # plot same but different color for each drv group
  geom_point() + 
  geom_smooth(se = F) # no conf. intrvl.
```

plot different point colors for each drv group 

```{r}
ggplot( data = mpg, mapping = aes( x = displ, y = hwy) ) + 
  geom_point(mapping = aes( color = drv)) +
  geom_smooth(se = F)
```

now plot different linetype for each drv group in smoothed line
```{r}
ggplot( data = mpg , mapping = aes( x = displ, y = hwy)) +
  geom_point(mapping = aes(color = drv)) +
  geom_smooth(mapping = aes(linetype = drv))
```




3.7 statistical transformations
```{r}
ggplot( data = diamonds ) + # plot bar chart of diamond dataset
  geom_bar(mapping = aes( x = cut))
```


reproduce barchart with the stat_count variable and not a geom
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

to override the default stat from count to something else use the following:
```{r}
demo = tribble( # create random numbers for diamond cut classes
  ~cut, ~freq, # x data, y data 
  "Fair", 1610,
  "Good", 4906,
  "Very Good", 12082,
  "Premium", 13791,
  "Ideal", 21551
)
ggplot(data = demo) + 
  geom_bar(mapping = aes( x = cut, y = freq), stat = "identity")
```



now override default mapping from transformed variable to aesthetics. Display bar chart of proportion rahter than count
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes( x = cut, y = stat(prop), group = 1)) # y variable transforms data to a proportion of 1 
```

for more computed variables see ?geom_bar -> computed variables

create summary plot to summarize y values for each unique x value
```{r}
ggplot( data = diamonds) +
  stat_summary(
    mapping = aes( x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
)
```


3.7.1 Exercises
Q. 1) what is the default geom
A.



3.8 Position adjustments


